---
title: "03-dplyr"
author: "Hannah E. Houts"
date: "`r Sys.Date()`"
output: html_document
---

# STAMPS 2023
## Lesson 3 - dplyr and Tidyverse

Tidyverse is a suite of powerful and cohesive R packages, including dplyr, ggplot2, and tidyr. These tools share similar syntax, and some people find them to be more intuitive. 

There are other benefits to using tidyverse, very nicely described in this [Carpentries lesson](https://datacarpentry.org/R-ecology-lesson/03-dplyr.html#what-are-dplyr-and-tidyr)

dplyr is an R package in the Tidyverse suite. Dplyr has commands that help address common data manipulation challenges:

`select()` picks variables (think columns) based on their names.
`filter()` picks cases based on their values (think filtering rows).
`arrange()` changes the ordering of the rows.
`relocate()` changes the order of columns.
`mutate()` adds new variables that are functions of existing variables
`summarise()` reduces multiple values down to a single summary.

These also combine with `group_by()` which allows you to perform any operation “by group”. 


## Loading in data for dplyr
Before we can read in our data in a dplyr-friendly way, we need to install some packages. 

Optionally, you can individually install and load readr, dplyr, ect like this
```{r}
#install.packages("readr")
#library(readr)

#install.packages("dplyr")
#library(dplyr)
```

but I prefer to just load the whole tidyverse suite like this:
```{r}
#install.packages("tidyverse")
library(tidyverse)
```

(its perfectly valid to only install individual packages, especially if you want to increase loading speed for a polished workflow or if you have a weird command conflict)

## readr

recall that we loaded in our tables using `read.csv`
We're going to add "baseR" to these variables
```{r}
covariates_baseR <- read.csv("data/FWS_covariates.txt", sep = "\t")
abundances_baseR <- read.csv("data/FWS_OTUs.txt", sep = "\t", row.names = 1, header = T)
```

When planning to manipulate data with dplyr or other Tidyverse tools, `read_csv()` or `read_tsv()` are preferable. 
`read_tsv()` is a readr command (part of the tidyverse)

```{r}
covariates <- read_tsv("data/FWS_covariates.txt")
abundances <- read_tsv("data/FWS_OTUs.txt")
```


#### Exercise: Please take a minute to explore the differences between the tables loaded with read.csv and read_tsv. Write down what you notice
- 


## select & filter

remind yourself what the table looks like:


```{r}
covariates
```


the dplyr command `select()` lets you select specific columns. 
```{r}
select(abundances, JPA_May, JPA_Jun)
```


piping lets you state the first argument (the name of the table) before the command, and "pipe in" the data like this:
```{r}
abundances %>% 
  select(JPA_May, JPA_Jun)
```

you can select with contains() or starts_with(), 

```{r}
abundances %>% 
  select(contains("Jan"))
```
you can also combine multiple select criteria
```{r}
abundances %>% 
  select(starts_with("LOP"))
```

```{r}
abundances %>% 
  select(!starts_with("MBL") & !contains("Jun"))
```

`filter()` is used to filter rows by their contents. 

remember this table
```{r}
covariates
```

```{r}
covariates %>% 
  filter(Location=="MBL")
```
you can combine filters too:
```{r}
covariates %>% 
  filter(Location=="MBL" & Season == "Winter") 
```


```{r}
names(abundances)
abundances %>% 
  filter(
    JPA_Jan == 0 &
    JPA_Feb == 0 &  
    JPA_May == 0 &  
    JPA_Jun == 0 & 
    LOP_Jan == 0 &  
    LOP_Feb == 0 &  
    LOP_Jun == 0 &  
    LOP_Jul == 0 &  
    MBL_Jan == 0 & 
    MBL_Feb == 0 &
    MBL_Jun == 0 &
    MBL_Aug == 0
    )
```
there are other ways to do this ;)

#### Exercise:

- Combine Select and Filter, generate a table from covariates that contains SampleName and Location, but only in the Summer season. 

- does it matter if you use select or filter first?




##  rename & relocate


notice that the first column has a weird name, probably because in the tsv there wasnt a column name, or these were "rownames" (rownames are not really used in tidyverse)
```{r}
names(abundances)
```


we can rename our "..1" column like this 
(note, the back ticks are required because the column name starts with a not-letter.)
```{r}
abundances <- abundances %>% 
  rename(taxonomy = `...1`)
abundances
```

we can move around column with `relocate()`. this is helpful if there is a particular few columns you are focusing on when exploring your data. 

if you want to move a column from the first position to the last one, try this
```{r}
abundances %>% 
  relocate(!taxonomy)
```


## mutate & arrange
mutate() creates a modified data frame with the newly added or modified columns based on your specifications.


```{r}
abundances %>%
  mutate(log_JPA_Jan = log(JPA_Jan)) %>% 
  relocate(JPA_Jan, log_JPA_Jan)
```


##

## group_by

## mutate redux
