---
title: "05: phyloseq (solutions)"
author: "Sarah Teichman"
date: "2023-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Phyloseq

`phyloseq` is an R package that is used to store, analyze, and plot microbiome abundance data. We will start by importing it from the Bioconductor package repository. 

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

if (!requireNamespace("phyloseq", quietly = TRUE)) {
     BiocManager::install("phyloseq")
}

# load in the phyloseq package
library(phyloseq)

# load in tidyverse
library(tidyverse)
```

If you have any issues installing `phyloseq` on your local computer, we recommend switching over to your R Studio server account where `phyloseq` is already installed.

## Load in data 

We'll be using the same `covariates` and `abundances` datasets and show how to make them into a `phyloseq` object. 

```{r}
covariates <- read.csv("data/FWS_covariates.txt", sep = "\t")
abundances <- read.csv("data/FWS_OTUs.txt", sep = "\t", row.names = 1, header = T)
```

## Creating phyloseq objects

A `phyloseq` object is made up of three tables. 

1. OTU table
2. Taxonomy table
3. Sample data table

Let's prepare these three tables from our data. 

```{r}
# our covariates data frame is our sample data table. 
# let's move the "SampleName" column to be row names for our data frame
row.names(covariates) <- covariates$SampleName
covariates <- covariates %>% select(-SampleName)

# our abundance table doesn't need to have all of the taxonomic information. 
# let's update the row names by numbering our OTUs. 
# first we'll save the current rownames, because we need that taxonomy information later
taxonomy <- row.names(abundances)
# replace row names with "taxa1" up through "taxa387" with the "paste0()" function
row.names(abundances) <- paste0("taxa", 1:nrow(abundances))

# now we can make a taxonomy table
# we would like to separate each taxonomic identification into separate columns for each hierarchical level 
# make a data frame for taxonomy 
tax_df <- data.frame(tax = taxonomy)
# separate the single column tax by creating a new column each time a ";" is seen
# name the newly created columns based on their rank 
tax_df_sep <- separate_wider_delim(data = tax_df, cols = tax, delim = ";",
                            names = c("kingdom", "phylum", "class",
                                      "order", "family", "genus"))
# turn the data frame into a matrix 
# (phyloseq prefers a matrix of taxonomy to a data frame)
# (I don't know why)
tax_mat <- as.matrix(tax_df_sep)
# make sure the row names of the taxonomy matrix match the row names of the OTU table 
row.names(tax_mat) <- row.names(abundances)
# now we can finally make a phyloseq object! 
phylo_obj <- phyloseq(otu_table(abundances, taxa_are_rows = TRUE), 
                      tax_table(tax_mat), 
                      sample_data(covariates))
phylo_obj
```

Great, we have a `phyloseq` object! But what have we actually done here? 

We have linked all of our information about our data (covariates for each sample, abundances, and taxonomic information about OTUs) into a single object. That's pretty cool! 

## Using phyloseq objects

Once we have our `phyloseq` object, we can do a lot with it! 

Let's start out by learning the basics of it (which we probably already know from our explorations of the data). 

```{r}
# check out our sample data 
nsamples(phylo_obj)
sample_names(phylo_obj)
sample_variables(phylo_obj)

# check out our abundance data 
sample_sums(phylo_obj)
ntaxa(phylo_obj)
head(taxa_sums(phylo_obj), 20)

# check out our taxonomy table
rank_names(phylo_obj)
```

We can also change the level of taxonomic classification we are working with. What if we only cared about differences up to the order level. 

```{r}
# use tax_glom to aggregate, where the second argument is the taxonomic level to aggregate to 
phylo_order <- tax_glom(phylo_obj, "order")
ntaxa(phylo_order)
head(tax_table(phylo_order))
# note that now the family and genus levels are all NA value because we have 
# aggregated to the phylum level 
# NA is a data type in R that refers to missing data or data that doesn't exist
```

We can also subset taxa. What if we only want to look at taxa from the phylum *Proteobacteria*? 

```{r}
# use subset_taxa to subset, using the second argument to specify which taxa you want at which taxonomic rank
phylo_proteo <- subset_taxa(phylo_obj, phylum == "Proteobacteria")
ntaxa(phylo_proteo)
head(tax_table(phylo_proteo))
```

## Plotting phyloseq objects

One very useful aspect of `phyloseq` is its analysis and plotting capacity. Although you could make any plot you want with the two data sets and some time spent with `dplyr` and `ggplot`, `phyloseq` makes this process more convenient for you. 

Imagine we want to look into alpha diversity for our data. 

```{r}
# estimate several types of alpha diversity 
estimate_richness(phylo_obj)
# plot Shannon and Simpson diversity measures by location 
# note: this plotting function is a wrapper for ggplot, so we can customize it as we would a ggplot object
plot_richness(phylo_obj, x = "Location", color = "Season", measures = c("Shannon", "Simpson")) + 
  # use a black and white background
  theme_bw() + 
  # add a title
  ggtitle("Shannon and Simpson diversities by location") + 
  # center the title
  theme(plot.title = element_text(hjust = 0.5))
```

Now imagine we want to make an ordination plot where we compare communities based on Bray-Curtis dissimilarities. 

```{r}
# calculate bray-curtis dissimilarities between each sample 
# then perform dimension reduction with nMDS (non-metric Multi Dimensional Scaling)
phylo_bray <- ordinate(phylo_obj, "NMDS", "bray")
# color our plot by season 
plot_ordination(phylo_obj, phylo_bray, type="samples", color="Season") +
  # use a black and white background
  theme_bw() + 
  # add a title
  ggtitle("nMDS of samples with Bray-Curtis dissimilarities") + 
  # center the title
  theme(plot.title = element_text(hjust = 0.5))
```

## Getting data out of phyloseq

Finally, there are some settings where you might rather work with individual data tables than with your phyloseq object. Pretend for a moment that you started with your `phylo_obj` object and have no access to the `covariates` or `abundances` data frames. `phyloseq` has handy functions for us to extract each of its three tables. 

```{r}
# extract sample data table
samp_tab <- sample_data(phylo_obj)
head(samp_tab)

# extract otu table
otu_tab <- otu_table(phylo_obj)
head(otu_tab)

# extract taxonomy table
taxon_tab <- tax_table(phylo_obj)
head(taxon_tab)
```

Now you have each of these tables as their own R objects and can manipulate and analyze them however you would like!

## Exercise

Using the `phylo_obj` object, subset to only taxa that belong to *Firmicutes*. Build an ordination plot based on the subsetted `phyloseq` object you've created using Bray-Curtis dissimilarities and nMDS (as in the above section) and color the points by the location where they were collected.

*Hint*: subset to a specific taxa with `subset_taxa()`. 

*Hint*: locations are saved as the "Location" variable in the `covariates` dataset.

```{r}
phylo_firm <- subset_taxa(phylo_obj, phylum == "Firmicutes")
phylo_firm_bray <- ordinate(phylo_firm, "NMDS", "bray")
# color our plot by season 
plot_ordination(phylo_firm, phylo_firm_bray, type="samples",
                color="Location") +
  # use a black and white background
  theme_bw() + 
  # add a title
  ggtitle("nMDS of samples with Bray-Curtis dissimilarities") + 
  # add a subtitle
  labs(subtitle = "only Firmicutes taxa") + 
  # center the title and subtitle
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

## References 

```{r}
citation("phyloseq")
```

Some materials adapted from this [tutorial](https://www.nicholas-ollberding.com/post/introduction-to-phyloseq/) by Nicholas Ollberding.