---
title: '06: ggplot'
author: "Hannah Houts & Bryshal Moore"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In yesterday's tutorial, we tested out generating a PCoA plot using ggplot (and vegan). Today, we'll delve into ggplot in more detail, exploring its various features and functionalities. 
 
ggplot2 (commonly just referred to as just ggplot) is a R package used to make high quality plots for your data. 
*ggplot2 is part of the tidyverse suite!*
It offers consistent syntax and allows for detailed customization of almost every element of your plots. Visualization options range from bargraphs, boxplots, scatterplot, pie charts, histograms and more. 

to get started, lets load tidyverse
(we're going to use a few dplyr functions in addition to ggplot, so loading tidyverse is a quick solution)
```{r}
library(tidyverse)
```

Lets check out a new science (fiction) dataset, and check out it's columns:
```{r}
powers <- read_csv("https://raw.githubusercontent.com/mblstamps/stamps2023/main/day1/intro_R/super_data/super_hero_powers.csv")
powers
```


see how a few of the names have spaces in them, and some of them start with uppercase letters? the package `janitor` has a function that "cleans up" column names according to their standards (read more here:https://rpubs.com/jenrichmond/clean_names)

```{r}
if (!requireNamespace("janitor", quietly = TRUE)) {
     install.packages("janitor")
}
library(janitor)

powers <- clean_names(powers)
powers
```

### Practice 1.
Try out janitor on the supers_hero_powers data
```{r}
superhero_info <- read_csv("https://raw.githubusercontent.com/mblstamps/stamps2023/main/day1/intro_R/super_data/heroes_information.csv")
# add code to clean names below!

superhero_info <- clean_names(superhero_info)
```



lets take that superpowers dataframe and make a plot!
```{r}
tabyl(dat = superhero_info$skin_color)

superhero_info %>% 
  arrange(desc(height))
```

say we want to make a bar chart to visualize the relative heights of superheros.
This is about the simplest version we can make using ggplot. 

```{r}
superhero_info %>% 
  ggplot(aes(x = name, y = height)) +
  geom_col()
```


Although the above code is equivalent to:
```{r}
  ggplot(superhero_info, aes(x = name, y = height)) +
  geom_col()
```

We tend to put the data first and then use the pipe (%>%) to send it to the ggplot() function. This is nice because when we add further data wrangling functions between the data and the ggplot(). For example, a filter!

```{r}
superhero_info %>% 
  filter(skin_color == "green") %>%
  ggplot(aes(x = name, y = height)) +
  geom_col()
```

The lines that come before the ggplot() function are piped, whereas from ggplot() onwards you have to use +. This is because we are now adding different layers and customization to the same plot.

aes() stands for aesthetics - things we can see. These set the axes for the plot, and infact you can just plot the aesthetics like this:
```{r}
superhero_info %>%
  filter(skin_color == "green") %>% 
  ggplot(aes(x = name, y = height)) 
```
Lets clean this up a little bit- we can fix those axis labels like this:
```{r}
superhero_info %>% 
  filter(skin_color == "green") %>%
  ggplot(aes(x = name, y = height)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ # hjust sets the alignment of the text.
  geom_col()
```
```{r}
superhero_info %>% 
  filter(skin_color == "green") %>%
  ggplot(aes(x = name, y = height)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ 
  labs(title = "Heights of Super Heros", x = "Hero Name", y = "Height (cm)")
```

Adding another aesthetic, fill, lets us fill in color by a variable of interest. 
```{r}
superhero_info %>% 
  filter(skin_color == "green") %>%
  ggplot(aes(x = name, y = height, fill = gender)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ 
  labs(title = "Heights of Super Heros", x = "Hero Name", y = "Height (cm)")
```


### Practice 2. 
edit the code below to make a barchart of your own, choosing your own:
- filter
- x, y, and fill aesthetic
- title and axis labels

try choosing a different x aesthetic in particular

(tip: for a barchart, usually y is numeric data. )
```{r}
superhero_info %>% 
  filter(publisher == "Marvel Comics" & alignment == "neutral") %>%
  ggplot(aes(x = gender, y = height)) +
  geom_col()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ 
  labs(title = "Heights of Super Heros", x = "Hero Name", y = "Height (cm)")
```

lets test out a scatter plot, by choosing a continuous variable for both the weights and height. note that the aesthetic for coloring the points is "color", not "fill", and the geom() layer is "point".

```{r}
superhero_info %>% 
  ggplot(aes(x = weight, y = height, color = gender)) +
  geom_point()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+ 
  labs(title = "Heights of Super Heros", x = "Weight", y = "Height (cm)")
```

## Applying to our microbial abundance data

Lets test out some plotting with our microbial abundance data, first by generating a relative abundance column by running the code below
```{r}
abundances <- read_tsv("data/FWS_OTUs.txt")

rel_abundances <- abundances %>%
    rename("taxonomy" = `...1`) %>%
    pivot_longer(!taxonomy, names_to = "name", values_to = "value") %>%
    separate(name, c("where", "when"), remove = F) %>%
    group_by(where, when) %>%
    mutate(relative_abundance = value/sum(value))
rel_abundances
```

Here we can make a scatter plot, plotting the relative abundance of each species by sample. adding `aplha = .5` in the geom layer makes the points 50% translucent
```{r}
rel_abundances %>% 
  ggplot(aes(x = name, 
             y = relative_abundance, 
             col = when)) + 
  geom_point(alpha=.5) +
  labs(x = "", y = "Relative\nabundance")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```



Lets test out a new plot. Here is code to make a relative abundances


```{r}
rel_abundances <- rel_abundances %>% mutate(when = factor(when, 
                       levels = c("Jan", "Feb", "May", "Jun", "Jul", "Aug"))) 


```

```{r}
rel_abundances %>% 
  filter(taxonomy == "Bacteria;Proteobacteria;Betaproteobacteria;Burkholderiales;Burkholderiaceae;Ralstonia") %>% 
  ggplot(aes(x = name, 
             y = relative_abundance, 
             col = when)) + 
  geom_point() +
  labs(y = "Relative\nabundance")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```

### Practice 3. 
Here is code to pivot the data again. first run this
```{r}
rel_abundances_wide <- abundances %>%
  rename("taxonomy" = 1) %>%
  pivot_longer(-1) %>%
  mutate(relative_abundance = value/sum(value)) %>%
  filter(taxonomy %in% c("Bacteria;Cyanobacteria;Cyanobacteria;SubsectionIV;Unassigned;Anabaena", "Bacteria;Proteobacteria;Betaproteobacteria;Burkholderiales;Burkholderiaceae;Ralstonia"))%>% 
  select(!value) %>%
  pivot_wider(names_from = taxonomy, 
              values_from = relative_abundance) %>%
  rename("Cyanobacteria" = 2, 
         "Proteobacteria" = 3)

rel_abundances_wide
```





Now, try to make a scatter plot of the relative abundances of Amy's favorite 2 taxa, against each other.

- add the x & y aesthetics to match the 2 columns in rel_abundances_wide that represent Amy's favorite taxa.
- debug the geom layer
- fill in the x & y labels, and add a title



```{r}
rel_abundances_wide %>% 
ggplot(aes(x =  Cyanobacteria,y = Proteobacteria)) + 
  geom_point() +
    labs(x="Amy's fav Cyanobacteria abundance", y = "Amy's fav Proteobacteria abundance", title = "Amy's fav Cyanobacteria abundance v. Amy's fav Proteobacteria abundance")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))
```


