---
title: '07: more on functions'
author: "Hannah Houts"
output: html_document
---

When are functions used in R?
 
Quite literally all the time :) 
 
when we read a csv, we are using a function, `read_csv()` from the package `readr`.

we typically use 
```
library(readr)
```
to load a package, 

but we can even load just one function from a package with this syntax:
```{r}
heros <- readr::read_csv("super_data/heroes_information.csv")
heros
```
 
 
```{r}
calc_count_by_eyecolor <- function(color){
  count <- sum(heros$`Eye color`==color)
  return(count)
}

calc_count_by_eyecolor("yellow")
```

```{r}
calc_avg_height_by_align <- function(align){
  align_df <- heros[heros$Alignment == align, ]
  count <- mean(align_df$Height)
  return(count)
}

calc_avg_height_by_align("good")
calc_avg_height_by_align("bad")

```
Challenge: make a function to calculate avg height by publisher

```{r}
calc_avg_height_by_pub <- function(align){
  align_df <- heros[heros$Publisher == align, ]
  count <- mean(align_df$Height, na.rm = T)
  return(count)
}
calc_avg_height_by_pub("George Lucas")

```


This is a functions to calculate the center log ratio of a vector of counts
```{r}
abundances <- read_tsv("data/FWS_OTUs.txt")

clr <- function(counts) {    
  log(counts) - mean(log(counts))
}

clr(c(5, 6, 7))
clr(c(5, 6, 0, 7))
```


```{r}
clr_better <- function(counts, psuedo = 0.5) {
  counts[counts == 0] <- psuedo
  log(counts) - mean(log(counts))
}
```


```{r}
clr_better(c(5, 6, 7))
clr_better(c(5, 6, 0, 7))
```


```{r}
abundances$LOP_Jan
```


```{r}
clr_better(abundances$LOP_Jan)
clr_better(abundances$LOP_Jan, psuedo = 0.1)

```




